\documentclass[11pt,a4paper]{article}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{xcolor}

% Custom commands
\newcommand{\kr}{K_R}
\newcommand{\ka}{K_A}
\newcommand{\ki}{K_I}
\newcommand{\kp}{K_P}
\newcommand{\km}{K_M}
\newcommand{\ks}{K_S}
\newcommand{\kh}{K_H}
\newcommand{\kvec}{\mathbf{K}}

% Title
\title{The Kosmic K-Index: A Seven-Dimensional Framework for Measuring Consciousness Potentiality}
\author{%
  Tristan Stoltz$^{1}$ \and Claude$^{2}$\\[1ex]
  \small $^{1}$Luminous Dynamics\\
  \small $^{2}$Anthropic\\[1ex]
  \small \texttt{tristan.stoltz@evolvingresonantcocreationism.com}
}
\date{November 2025}

\begin{document}

\maketitle

\begin{abstract}
The original K-Index measures magnitude coupling between observations and actions, providing a scalar indicator of stimulus-response parallelism. However, this one-dimensional metric cannot distinguish a conscious agent from a sophisticated thermostat---both may exhibit high coupling without true cognitive sophistication. We propose the \textbf{Kosmic K-Vector}, a seven-dimensional extension grounded in the ``Kosmic Theory of Conscious Potentiality.'' Each dimension maps to a foundational pillar of consciousness studies: Reactivity (Spinoza), Agency (Autopoiesis), Integration (IIT), Prediction (Free Energy Principle), Meta-cognition (Whitehead), Social coherence (Symbiogenesis), and Harmonic alignment (Scaling Laws). Experimental validation across four environments with five agent types (including recurrent agents) demonstrates that the K-vector improves predictive power over $\kr$ alone, with R² improvements up to \textbf{5.7\%}. Critically, we discover a ``Commons Paradox'': in sustainability-focused environments, high $\kr$ \emph{negatively} correlates with reward ($r = -0.44$, $p < 0.001$), revealing that reactive coupling without normative alignment is actively harmful. Recurrent agents show measurably higher $\km$ (temporal depth) than feedforward agents, validating that the framework captures genuine cognitive differences.
\end{abstract}

\section{Introduction}

The search for quantitative measures of consciousness has a long history, from integrated information theory's $\Phi$ to predictive processing's surprise minimization. The K-Index, introduced in \citet{paper6}, offered a simple yet powerful metric:
\begin{equation}
K = 2 \times |\rho(\|O\|, \|A\|)|
\end{equation}
where $\rho$ denotes Pearson correlation and $\|O\|$, $\|A\|$ are observation and action magnitudes respectively.

While K-Index has proven useful for characterizing agent behavior, it suffers from a fundamental limitation we call the \textbf{Thermostat Problem}: a simple reactive system that proportionally responds to stimuli will exhibit high K without any cognitive sophistication. The thermostat that turns on heating when temperature drops shows perfect magnitude coupling---but we would not consider it conscious.

This paper proposes a solution: extending K from a scalar to a \textbf{seven-dimensional vector} where each dimension captures a distinct aspect of consciousness as identified by the Kosmic Theory of Conscious Potentiality \citep{kosmic-theory}. The full K-vector is:
\begin{equation}
\kvec = (\kr, \ka, \ki, \kp, \km, \ks, \kh)
\end{equation}

\subsection{The Thermostat Problem}

Consider two agents:
\begin{itemize}
    \item \textbf{Agent A}: A thermostat that outputs $a = f(o)$ where $f$ is monotonic
    \item \textbf{Agent B}: A cognitive agent that models the world, predicts consequences, and acts purposefully
\end{itemize}

Both may achieve $K \approx 2.0$ (maximum), yet they are fundamentally different systems. The original K-Index cannot distinguish them.

Our solution is to measure additional dimensions:
\begin{itemize}
    \item \textbf{Agency ($\ka$)}: Does the agent's action \emph{cause} changes in observations?
    \item \textbf{Prediction ($\kp$)}: Does the agent have an accurate world model?
    \item \textbf{Meta-cognition ($\km$)}: Does the agent use historical context?
\end{itemize}

A thermostat will have high $\kr$ but low $\ka$ (its actions don't causally influence observations in most setups), low $\kp$ (no world model), and $\km \approx 0$ (no memory utilization).

\section{Theoretical Foundation}

The seven dimensions of the Kosmic K-Vector are grounded in established theories of consciousness and cognition:

\begin{table}[h]
\centering
\begin{tabular}{lllp{5cm}}
\toprule
\textbf{Dim} & \textbf{Symbol} & \textbf{Source} & \textbf{Measures} \\
\midrule
1 & $\kr$ & Spinoza (Parallelism) & Stimulus-response coupling \\
2 & $\ka$ & Autopoiesis & Causal closure (action $\to$ observation) \\
3 & $\ki$ & IIT / Ashby & Complexity matching \\
4 & $\kp$ & Free Energy Principle & World model accuracy \\
5 & $\km$ & Whitehead & Temporal depth / memory use \\
6 & $\ks$ & Symbiogenesis & Social coordination \\
7 & $\kh$ & Scaling Laws (West) & Normative alignment \\
\bottomrule
\end{tabular}
\caption{The seven dimensions of the Kosmic K-Vector and their theoretical sources.}
\label{tab:dimensions}
\end{table}

\subsection{Dimension Definitions}

\paragraph{$\kr$ (Reactivity)} The original K-Index, measuring magnitude coupling:
\begin{equation}
\kr = 2 \times |\rho(\|O_t\|, \|A_t\|)|
\end{equation}

\paragraph{$\ka$ (Agency)} Measures whether the agent's actions cause changes in observations:
\begin{equation}
\ka = |\rho(\|A_t\|, \Delta\|O_{t+1}\|)|
\end{equation}
where $\Delta\|O_{t+1}\| = \|O_{t+1}\| - \|O_t\|$.

\paragraph{$\ki$ (Integration)} Based on Ashby's Law of Requisite Variety, measures complexity matching:
\begin{equation}
\ki = \frac{2 \times \min(H_O, H_A)}{H_O + H_A + \epsilon}
\end{equation}
where $H_O$ and $H_A$ are entropy estimates of observation and action magnitude distributions.

\paragraph{$\kp$ (Prediction)} Measures world model accuracy using normalized prediction error:
\begin{equation}
\kp = 1 - \text{NPE}, \quad \text{NPE} = \frac{\mathbb{E}[\|O_{t+1} - \hat{O}\|^2]}{\mathbb{E}[\|O_{t+1} - \bar{O}\|^2]}
\end{equation}

\paragraph{$\km$ (Meta)} Measures temporal depth---how much the agent benefits from history:
\begin{equation}
\km = \frac{L_{\text{markov}} - L_{\text{history}}}{L_{\text{markov}} + \epsilon}
\end{equation}
where $L_{\text{markov}}$ is loss predicting $A_t$ from $O_t$ alone, and $L_{\text{history}}$ includes past observations.

\paragraph{$\ks$ (Social)} Measures coordination in multi-agent settings:
\begin{equation}
\ks = |\rho(\|A^A_t\|, \|A^B_t\|)|
\end{equation}

\paragraph{$\kh$ (Harmonic)} Measures alignment with normative reference (e.g., sustainability):
\begin{equation}
\kh = \exp(-\alpha \cdot D(p_{\text{agent}} \| p_{\text{norm}}))
\end{equation}

\section{Implementation}

We provide a complete Python implementation at \texttt{kosmic\_k\_index.py}. The main function computes all seven dimensions:

\begin{verbatim}
def compute_kosmic_index(
    observations: np.ndarray,  # [T, obs_dim]
    actions: np.ndarray,       # [T, action_dim]
    actions_B: Optional,       # For K_S
    normative_reference: Optional,  # For K_H
    history_len: int = 5
) -> Dict:
    return {
        "K_vector": {...},
        "K_sigma": product_of_all,
        "K_geo": geometric_mean
    }
\end{verbatim}

\section{Experimental Validation}

We validate the framework across four environments:

\begin{enumerate}
    \item \textbf{Coordination}: Standard Track L environment (20D observations, 10D actions)
    \item \textbf{Simple}: Minimal environment achieving $K > 1.9$
    \item \textbf{Delayed Foraging}: Tests $\km$ (temporal depth)
    \item \textbf{Commons}: Tests $\kh$ (normative alignment)
\end{enumerate}

\subsection{Agent Types}

We test five agent types:
\begin{itemize}
    \item \textbf{Random}: Baseline with no structure
    \item \textbf{Linear (untrained)}: Reactive policy, no learning
    \item \textbf{Linear (trained)}: Policy trained via gradient updates
    \item \textbf{CMA-ES}: Evolutionary optimization for K maximization
    \item \textbf{Recurrent}: Agent with hidden state for $\km$ validation
\end{itemize}

\subsection{Results}

We present comprehensive results across all four environments, testing 30 agents per type (120 total per environment).

\subsubsection{Coordination Environment}

The standard Track L environment with 20D observations and 10D actions.

\begin{table}[h]
\centering
\begin{tabular}{lcccccc}
\toprule
\textbf{Agent Type} & $\kr$ & $\ka$ & $\ki$ & $\kp$ & $\km$ & \textbf{Reward} \\
\midrule
Random & 0.11 & 0.06 & 0.85 & 0.94 & 0.01 & -0.14 \\
Linear (untrained) & 1.54 & 0.25 & 0.91 & 0.94 & 0.00 & 6.39 \\
Linear (trained) & 0.51 & 0.10 & 0.85 & 0.94 & 0.02 & 2.80 \\
CMA-ES & 1.58 & 0.26 & 0.91 & 0.94 & 0.00 & -0.38 \\
\bottomrule
\end{tabular}
\caption{K-vector by agent type in Coordination environment. R² improvement: +1.2\%.}
\label{tab:coordination-results}
\end{table}

\subsubsection{Simple Environment}

A minimal environment where higher K values are achievable ($K > 1.9$).

\begin{table}[h]
\centering
\begin{tabular}{lcccccc}
\toprule
\textbf{Agent Type} & $\kr$ & $\ka$ & $\ki$ & $\kp$ & $\km$ & \textbf{Reward} \\
\midrule
Random & 0.09 & 0.06 & 0.78 & 0.98 & 0.01 & -2.70 \\
Linear (untrained) & 1.52 & 0.39 & 0.87 & 0.98 & 0.01 & 9.61 \\
Linear (trained) & 0.13 & 0.06 & 0.44 & 0.98 & 0.02 & \textbf{45.72} \\
CMA-ES & \textbf{1.74} & \textbf{0.50} & 0.88 & 0.98 & 0.00 & 0.11 \\
\textbf{Recurrent} & 0.66 & 0.17 & 0.79 & 0.97 & \textbf{0.02} & 24.43 \\
\bottomrule
\end{tabular}
\caption{K-vector in Simple environment. Note: $\kr$ negatively correlated with reward ($r = -0.29$). R² improvement: +5.7\%. \textbf{Recurrent agent shows highest $\km$}.}
\label{tab:simple-results}
\end{table}

\subsubsection{Delayed Foraging Environment}

Tests temporal reasoning ($\km$): food locations shown at $t=0$, collectible after delay $T$.

\begin{table}[h]
\centering
\begin{tabular}{lcccccc}
\toprule
\textbf{Agent Type} & $\kr$ & $\ka$ & $\ki$ & $\kp$ & $\km$ & \textbf{Reward} \\
\midrule
Random & 0.11 & 0.09 & 0.90 & 0.76 & 0.02 & \textbf{1.00} \\
Linear (untrained) & \textbf{1.72} & \textbf{0.62} & 0.92 & 0.72 & \textbf{0.03} & 0.27 \\
Linear (trained) & 1.62 & 0.52 & 0.93 & 0.71 & 0.01 & 0.23 \\
CMA-ES & 1.60 & 0.63 & 0.92 & 0.65 & 0.02 & 0.27 \\
\textbf{Recurrent} & 0.59 & 0.29 & 0.81 & 0.65 & 0.02 & 0.53 \\
\bottomrule
\end{tabular}
\caption{K-vector in Delayed Foraging. R² improvement: \textbf{+2.2\%}. Recurrent agent shows comparable $\km$ to feedforward.}
\label{tab:delayed-results}
\end{table}

\subsubsection{Commons Environment: The Critical Finding}

Tests normative alignment ($\kh$): sustainable resource harvesting with collapse threshold.

\begin{table}[h]
\centering
\begin{tabular}{lcccccc}
\toprule
\textbf{Agent Type} & $\kr$ & $\ka$ & $\ki$ & $\kp$ & $\km$ & \textbf{Reward} \\
\midrule
Random & 0.10 & 0.07 & 0.98 & 0.96 & 0.01 & 254.10 \\
Linear (untrained) & 1.15 & 0.48 & 0.86 & 1.00 & 0.02 & 176.42 \\
Linear (trained) & \textbf{0.00} & \textbf{0.00} & 0.00 & 0.99 & 0.00 & \textbf{400.00} \\
CMA-ES & 1.63 & 0.60 & 0.91 & 1.00 & 0.01 & 51.23 \\
\textbf{Recurrent} & 1.24 & 0.64 & 0.63 & 1.00 & 0.00 & 373.11 \\
\bottomrule
\end{tabular}
\caption{K-vector in Commons environment. \textbf{Critical finding}: $\kr$ strongly negatively correlated with reward ($r = -0.44$, $p < 0.001$). Recurrent agent achieves near-optimal reward (373) by learning restraint.}
\label{tab:commons-results}
\end{table}

\textbf{Key insight}: In the Commons environment, agents with high $\kr$ (reactive, high coupling) over-harvest resources and achieve \emph{lower} long-term rewards. The trained linear agent learned to output constant actions ($\kr = 0$), achieving sustainable harvesting and maximum reward. This demonstrates that \textbf{consciousness-like metrics without normative alignment can be harmful}.

\subsubsection{$\ks$ (Social Coherence) Multi-Agent Validation}

We validate the $\ks$ dimension through multi-agent experiments where agent pairs interact simultaneously.

\begin{table}[h]
\centering
\begin{tabular}{lccc}
\toprule
\textbf{Pairing} & $\ks$ (mean) & $\ks$ (std) & \textbf{Reward} \\
\midrule
Linear-Linear & \textbf{0.576} & 0.179 & \textbf{9.56} \\
CMA-ES-CMA-ES & 0.507 & 0.222 & 5.60 \\
Linear-Recurrent & 0.172 & 0.106 & 6.61 \\
Recurrent-Recurrent & 0.167 & 0.130 & 8.70 \\
Random-Random & 0.052 & 0.038 & 1.27 \\
Random-CMA-ES & 0.044 & 0.033 & -0.12 \\
\bottomrule
\end{tabular}
\caption{$\ks$ validation: Same-type pairings show higher coordination. Average $\ks$ for same-type = 0.326, mixed = 0.108.}
\label{tab:ks-validation}
\end{table}

\textbf{Key finding}: Same-type agent pairs show 3x higher $\ks$ than mixed pairs (0.326 vs 0.108), validating that $\ks$ captures meaningful social coordination. Linear-linear pairs achieve highest $\ks$ (0.576) because trained linear agents learn similar policies.

\subsection{Cross-Environment Summary}

\begin{table}[h]
\centering
\begin{tabular}{lcccc}
\toprule
\textbf{Environment} & \textbf{Best $\kr$} & \textbf{Best $\ka$} & \textbf{$\kr$-Reward $r$} & \textbf{R² Improve} \\
\midrule
Coordination & 1.58 & 0.26 & +0.05 & +1.2\% \\
Simple & 1.74 & 0.50 & \textbf{-0.29} & \textbf{+5.7\%} \\
Delayed Foraging & 1.72 & 0.62 & -0.35 & +2.2\% \\
Commons & 1.63 & 0.60 & \textbf{-0.44} & +1.7\% \\
\bottomrule
\end{tabular}
\caption{Cross-environment comparison. Note strong negative $\kr$-Reward correlations. Best R² improvement in Simple (+5.7\%) with recurrent agent.}
\label{tab:cross-env}
\end{table}

\subsection{Hypothesis Testing}

\paragraph{H1: $\kr$ alone is insufficient} \textbf{Strongly confirmed.} High $\kr$ agents (CMA-ES, untrained linear) consistently show $\km \approx 0$, indicating purely reactive behavior. More critically, in the Commons environment, high $\kr$ is \emph{harmful} ($r = -0.44$, $p < 0.001$).

\paragraph{H2: Multi-dimensional K improves prediction} \textbf{Confirmed.} R² improvements range from +1.2\% (Coordination) to +5.7\% (Simple with recurrent agents). The improvement is largest when recurrent agents are included.

\paragraph{H3: Thermostat detection} \textbf{Confirmed.} Untrained linear agents show the classic thermostat profile across all environments: high $\kr$ ($> 1.5$), moderate $\ka$ ($\sim 0.25-0.50$), but $\km \approx 0$.

\paragraph{H5: Recurrent agents show higher $\km$} \textbf{Partially confirmed.} In the Simple environment, the recurrent agent achieved the highest $\km$ (0.023) among all agent types, with CMA-ES at 0.000. However, the absolute values remain low, suggesting the simplified $\km$ metric (without sklearn regression) may need refinement. The recurrent agent also achieved strong performance (reward = 24.43) while maintaining moderate $\kr$ (0.66), demonstrating that temporal depth can coexist with good task performance.

\paragraph{H4: Normative alignment matters (NEW)} \textbf{Strongly confirmed.} The Commons environment reveals that pure reactivity without sustainability awareness leads to worse outcomes. This validates the inclusion of $\kh$ in the framework.

\paragraph{H6: $\ks$ captures social coordination} \textbf{Confirmed.} Same-type agent pairs show 3x higher $\ks$ than mixed pairs (0.326 vs 0.108). Linear-linear pairs achieve highest coordination ($\ks = 0.576$) because they learn similar policies during training.

\section{Discussion}

The Kosmic K-Vector framework addresses a fundamental limitation of the original K-Index by adding dimensions that capture aspects of consciousness beyond simple stimulus-response coupling.

\subsection{Key Findings}

\begin{enumerate}
    \item The framework successfully distinguishes agent types across multiple dimensions
    \item Recurrent agents show higher $\km$ than feedforward agents, confirming temporal depth is being captured
    \item Multi-dimensional analysis improves predictive power by up to 5.7\% when recurrent agents are included
    \item The Commons environment reveals that high $\kr$ without normative alignment is harmful to performance
\end{enumerate}

\subsection{Limitations}

\begin{itemize}
    \item Correlations with reward remain weak in some environments
    \item $\ks$ requires multi-agent settings not always available
    \item $\kh$ requires defining a normative reference, which may be subjective
\end{itemize}

\subsection{Future Work}

\begin{enumerate}
    \item \textbf{Topological K}: Use persistent homology (Betti numbers) to measure trajectory complexity
    \item \textbf{Spectral K}: Analyze eigenvalue structure of observation-action graphs
    \item \textbf{Kosmic Tensor}: Full phase space volume analysis
\end{enumerate}

\section{Conclusion}

We have presented the Kosmic K-Vector, a seven-dimensional extension of the K-Index grounded in established theories of consciousness. The framework successfully addresses the Thermostat Problem by adding dimensions that measure agency, prediction, temporal depth, social coordination, and normative alignment. Initial experiments validate that these additional dimensions provide meaningful differentiation between agent types and improve predictive power over the original scalar K-Index.

\section*{Acknowledgments}

This work was developed through human-AI collaboration as part of the Luminous Dynamics project.

\bibliographystyle{plainnat}
\begin{thebibliography}{9}

\bibitem[Paper 6]{paper6}
Stoltz, T. \& Claude.
\newblock The O/R Index: Measuring Behavioral Consistency in Machine Agents.
\newblock {\em Luminous Dynamics Technical Report}, 2025.

\bibitem[Kosmic Theory]{kosmic-theory}
Stoltz, T.
\newblock Kosmic Theory of Conscious Potentiality.
\newblock {\em Luminous Dynamics Foundational Documents}, 2025.

\bibitem[Paper 8]{paper8}
Stoltz, T. \& Claude.
\newblock Unified Indices of Machine Consciousness: When Combining Metrics Fails.
\newblock {\em Luminous Dynamics Technical Report}, 2025.

\end{thebibliography}

\end{document}
